---
title: "Energy Dataset CLeaning"
author: "Anurag Thakur"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---


## Required packages 
```{r}
#install.packages("readr")
library(readr)
#install the tidyr package
#install.packages("tidyr")
#load the tidyr package
 library(tidyr)
library(dplyr)
library(dplyr)
library(stringr)
```


## Executive Summary 

Data preprocessing is a must have tool in a data scientist's toolbox and I selected these data sets due to my interest in transcition from non-renewable to renewable energy for energy production.



## Data 

The datasets chosen were 2 data from kaggle.
1st Dataset :- This dataset contains energy statitics of production, trade, conversion and final consumption of different sources of energy.
Source :- https://www.kaggle.com/unitednations/international-energy-statistics
published by the United Nations Statistics Division

2nd Datasets :- This dataset contains observations renewable energy production from enviornment friendly sources. 
Source : https://www.kaggle.com/khadeejahalghadeer/renewable-energy-generation-world-1965-to-2018
published by http://www.bp.com/statisticalreview



```{r}
# This is the R chunk for the Data Section
Data1 <- read_csv("all_energy_statistics.csv")
Data2  <- read_csv("modern-renewable-energy-consumption.csv")
Data1.1 <- Data1 %>% select( - (2))
Data1.1 <- Data1.1 %>% select(- (5:6))
colnames(Data1.1)[1] = "Country"
colnames(Data1.1)[2] = "Year"

```
Explanation of relevant variables:-
DATASET 1
country_or_area :- country, year = timeline , unit  = unit of consuption, quantity  = consumption

DATASET 2
Entity = country, year = timeline, Hydropower / Solar /Wind /Other renewables - profuction in terawatt
## Understand 

Summarising the types of variables and data structures,

Checking the variable types of 1st DATASET

```{r}
print("Structure of 1st Dataset")
str(Data1)

class(Data1$country_or_area)
class(Data1$year)
class(Data1$quantity)
class(Data1$unit)

print("no change required")
```
Checking the variable types of 2st DATASET

```{r}

print("Structure of 2nd Dataset")
str(Data2)

print("no change required")
class(Data2$Entity)
class(Data2$Year)
class(Data2$`Hydropower (terawatt-hours)`)
class(Data2$`Solar (terawatt-hours)`)
class(Data2$`Wind (terawatt-hours)`)
class(Data2$`Other renewables (terawatt-hours)`)
print("no change required")



```

##	Tidy & Manipulate Data I 
Removed the unnecessary columns comodity transaction(column 2) and quantity foot notes(column 6) and catagory(column 7)
Renamed the columns

```{r}
#removing unnecessary columns data1

Data1.1 <- Data1 %>% select( - (2))
Data1.1 <- Data1.1 %>% select(- (5:6))
colnames(Data1.1)[1] = "Country"
colnames(Data1.1)[2] = "Year"
head(Data1.1,10)
```

removing unnecessary column - code (column 2)

```{r}
#removing unnecessary columns data2
Data2.2 <-  Data2 %>% select( - (2))

colnames(Data2.2)[1] = "Country"

head(Data2.2,10)
```


##Merging the datasets
Removing White spaces,
Filtering data to year 2014,
and removing time from 1st data set variable as we have already filtered the data,
Triming the down to distinct country valuesin both datasets.


```{r}

Data1.1 <- Data1.1 %>% mutate_if(is.character, str_trim)

DataDistict <-Data1.1 %>% filter(Year == 2014)
DataDistict <-DataDistict %>% distinct(Country, .keep_all = TRUE)
DataDistict <- DataDistict %>% select( -(2)) 
head(DataDistict,10)
```

```{r}
Data2.2 <- Data2.2 %>%  mutate_if(is.character, str_trim)

DataDistict_2 <-Data2.2 %>% filter(Year == 2014)
DataDistict_2<-DataDistict_2 %>% distinct(Country, .keep_all = TRUE)
head(DataDistict_2,10)

```

Merging the data on the basis of column country using right join i.e. prioritizing the observations of dataframe DATA2.1 i.e., table containing observations of produciotn of energy using renewbale resources.
```{r}
Final_data <- right_join(DataDistict,DataDistict_2, key = "Country")
head(Final_data,10)
```

##	Tidy & Manipulate Data II 

Creating 2 new variable from the existing variables (`Hydropower (terawatt-hours)` + `Solar (terawatt-hours)`+`Wind (terawatt-hours)`+`Other renewables (terawatt-hours)`).

Total_Renewable = sum of the energy produced from renewable resources
Hydropower_Percentage = contribution of Hydropower to total energy produced by renewable resources

Renaming the column

Removing the variables not required anymore `Hydropower (terawatt-hours)` + `Solar (terawatt-hours)`+`Wind (terawatt-hours)`+`Other renewables (terawatt-hours)`

```{r}
Final_data <- Final_data %>% mutate(
  Total_Renewable = `Hydropower (terawatt-hours)` + `Solar (terawatt-hours)`+`Wind (terawatt-hours)`+`Other renewables (terawatt-hours)`,
  Hydropower_Percentage = `Hydropower (terawatt-hours)`/ Total_Renewable *100)
Final_data$Total_Renewable %>% as.numeric()
Final_data <- Final_data %>% select( -(5:8) )
head(Final_data,10)

```

Diving the unit column into two
```{r}

Final_data %>% separate(unit, into = c("Unit", " Multiplicant"), sep = ",")
```

##	Scan I 

Scanning the numeric attributes for missing values, special values and obvious errors (i.e. inconsistencies).

```{r}
print("missing values in the entire dataset")
sum(is.na(Final_data))
print("missing values in the unit column")
sum(is.na(Final_data$unit))
print("missing values in the quantity column")
sum(is.na(Final_data$quantity))
print("missing values in the Total_Renewable column")
sum(is.na(Final_data$Total_Renewable))
print("missing values in the Hydropower_Percentage column")
sum(is.na(Final_data$Hydropower_Percentage))

```
Results for Inconsistency Operations for unit and multiplicant column are the same, so only one column is displayed
in unit column NA values cannot be treated
in quantity column NA values can be treated as the units are different, operations for conversion are irrelevant to this report
treating NA values for Total_Renewable and Hydropower_Percentage by replacing them with mean


```{r}

print("infinite values in the unit column")
sum(is.infinite(Final_data$unit))
print("infinite values in the quantity column")
sum(is.infinite(Final_data$quantity))
print("infinite values in the Total_Renewable column")
sum(is.infinite(Final_data$Total_Renewable))
print("infinite values in the Hydropower_Percentage column")
sum(is.infinite(Final_data$Hydropower_Percentage))

```

##	Scan II

Scanning the numeric data(Total renewable and Hydropower_Percentage) for outliers. 


```{r}
par(mfrow=c(1,3))
# This is the R chunk for the Scan II
Final_data$Total_Renewable %>%  boxplot(main="Box Plot of Total_Renewable", ylab="terawatt-hour", col = "yellow")

Final_data <-  Final_data %>%filter(Final_data$Total_Renewable < 1500)

Final_data$Total_Renewable %>%  boxplot(main="Box Plot of Total_Renewable", ylab="terawatt-hour", col = "green")

Final_data <-  Final_data %>%filter(Final_data$Total_Renewable < 400)

Final_data$Total_Renewable %>%  boxplot(main="Box Plot of Total_Renewable", ylab="terawatt-hour", col = "blue")

```

No outliers were found for the total
```{r}
# This is the R chunk for the Scan II
Final_data$Hydropower_Percentage %>%  boxplot(main="Box Plot of Percentage contribution of Hydropower to total renewable resources", ylab="terawatt-hour", col = "red")


```

##	Transform 


Transformation task was performed on 2 variables,

for Transformation the variable Total Renewable attribute, logarithmic transmission gave us better results
```{r}
par(mfrow=c(1,3))
hist(Final_data$Total_Renewable,col = blues9)
scale_x1 <- scale(Final_data$Total_Renewable, center = TRUE, scale = TRUE)
hist(scale_x1,col = rainbow(7) )
log_Total_Renewable <- log(Final_data$Total_Renewable)
hist(log_Total_Renewable, ,col = rainbow(7))
#hist(Final_data$quantity)
```
Unlike with the Total Renewable attribute, transformation using logarithms wasn't as effective as scaling.

```{r}
par(mfrow=c(1,3))
hist(Final_data$Hydropower_Percentage, col = blues9)
log_Total_Renewable <- log(Final_data$Hydropower_Percentage)
hist(log_Total_Renewable,col = rainbow(7))
scale_x1 <- scale(Final_data$Hydropower_Percentage, center = TRUE, scale = TRUE)
hist(scale_x1 , col = rainbow(4) )

#hist(Final_data$quantity)
```
Thank you!